---
title: "Weather and Corn Yield Regressions"
author: "Lauren Kremer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = FALSE,
                      fig.keep='all')
library(tidyverse)
library(R.matlab)
library(rnassqs)



#devtools::install_github('UrbanInstitute/urbnmapr')
library(urbnmapr)
library(USAboundaries) #USA states and counties
library(tidyr)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE,
                      fig.keep='all')
setwd('/Users/laurenkremer/Dropbox/PhD_coursework/ESS580_environmental_science/weather_corn_regressions')
# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r tidying up, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)
# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & doy <= 59 | doy >= 335 & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)

# set our API key with NASS
nassqs_auth(key = "C262F5FB-1538-3D7C-88C8-75A1C21A7767")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?
```{r}
# starting with cornyields df 
#names(cornyields)
#unique(cornyields$county_name)

winne_cy <- cornyields %>%
  filter(county_name=='WINNESHIEK'  & !is.na(yield))


ggplotRegression <- function (fit) {

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  labs(y = 'Corn yield (bushels/acre)', x = 'Year') +
  stat_smooth(method = "lm", col = "purple") +
  annotate('text', x = 2005, y = 100, label = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))
}

fit1 <- lm(yield ~ year, data = winne_cy)
ggplotRegression(fit1)
```

**RESPONSE** <br/>
**Yes, there is a significant increase in corn yield over time for Winneshiek County, IA**

### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 
```{r}
winne_cy$yearsq <- winne_cy$year^2

lm_winne_cyquad <- lm(yield ~ year + yearsq, winne_cy)
lm_cy <- summary(lm_winne_cyquad)

# adding squared term can see non-linearities added to time trend.
winne_cy$fitted <-lm_winne_cyquad$fitted.values

ggplot(winne_cy) +
  geom_point(aes(year, yield)) +
  geom_line(aes(year, fitted))+
  theme_bw() +
  labs(y = 'Corn yield (bushels/acre)', x = 'Year') 
```
**RESPONSE** <br/>
**No, there does not appear to be a change in yield growth over time for Winneshiek County, IA**

### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.
```{r}
str(winnesummer)
str(winne_cy)

cy_temps <- inner_join(winne_cy, winnesummer %>%
                            distinct(year,.keep_all=T),
                          by='year')

cy_temps$yearsq <- cy_temps$year^2
cy_temps$tmaxsq <- cy_temps$meantmax^2

lm_tmaxquad <- lm(yield ~ meantmax + tmaxsq, cy_temps)
summ_lm_tmax <- summary(lm_tmaxquad)

# adding squared term can see non-linearities added to time trend.
cy_temps$fitted <-lm_tmaxquad$fitted.values


ggplot(cy_temps) +
  geom_point(aes(meantmax, yield)) +
  geom_line(aes(meantmax, fitted))+
  theme_bw() +
  labs(x = 'Mean summer maximum temperature (°C)', x = 'Corn yield (bushels/acre)') 

```
**RESPONSE** <br/>
**Corn yield appears to decrease with increasing temperature in Winneshiek County, IA** <br/>

### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.
```{r}
tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

allia_summer <-tmaxdf %>%
  filter(year==2018 & doy >= 152 & doy <= 243) 

ia_yields_2018 <-cornyields %>%
  filter(year==2018) %>%
  mutate(countyfp = as.factor(county_ansi))

cy_temps <- left_join(ia_yields_2018, allia_summer %>%
                            distinct(countyfp,.keep_all=T),
                          by= c('countyfp', 'year'))


ggplotRegression <- function (fit) {

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  labs(y = 'Corn yield (bushels/acre)', x = 'Mean summer maximum temperature (°C)') +
  stat_smooth(method = "lm", col = "purple") +
  annotate('text', x = 31, y = 230, label = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))
}


fit1 <- lm(yield ~ tmax, data = cy_temps)
ggplotRegression(fit1)

```

**RESPONSE** <br/>
**There appears to be a significant decrease in corn yield with increasing temperature for all Iowa counties in this dataset (i.e. corn crops are less productive in hot summers)** <br/>

### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.
```{r}
# use cy_temps from above with includes a point for each county which is a factor column
str(cy_temps)

cy_temps$yearsq <- cy_temps$year^2
cy_temps$tmaxsq <- cy_temps$tmax^2

lm_tmaxquad <- lm(yield ~ tmax + tmaxsq + countyfp + year, cy_temps)
summ_lm_tmax <- summary(lm_tmaxquad)

# adding squared term can see non-linearities added to time trend.
cy_temps$fitted <-lm_tmaxquad$fitted.values

ggplot(cy_temps) +
  geom_point(aes(tmax, yield)) +
  geom_line(aes(tmax, fitted))+
  theme_bw() +
  labs(x = 'Mean summer maximum temperature (°C)', x = 'Corn yield (bushels/acre)') 

```
**RESPONSE** <br/>
**The county code is not a variable that influences the relationship between temperature and yield, rather it is an arbitrary factor variable. A linear model that includes this variable could obscure the relationships among other variables.  ** <br/>

### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.
```{r, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = FALSE)

nassqs_auth(key = "C262F5FB-1538-3D7C-88C8-75A1C21A7767")

# parameters to query on 
params <- list(commodity_desc = "SOYBEANS", statisticcat_desc = "YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1985, state_alpha = "IA")

# download
soyieldsall <- nassqs_yields(params)

soyieldsall$county_ansi <- as.numeric(soyieldsall$county_ansi)
soyieldsall$yield <- as.numeric(soyieldsall$Value)

# clean and filter this dataset
soyields <- select(soyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyields <- tibble(soyields)
```
```{r}
tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

allia_summer <-tmaxdf %>%
  filter(year==2018 & doy >= 152 & doy <= 243) 

ia_yields_2018 <-soyields %>%
  filter(year==2018) %>%
  mutate(countyfp = as.factor(county_ansi))

sy_temps <- left_join(ia_yields_2018, allia_summer %>%
                            distinct(countyfp,.keep_all=T),
                          by= c('countyfp', 'year'))


ggplotRegression <- function (fit) {

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  labs(y = 'Soy yield (bushels/acre)', x = 'Mean summer maximum temperature (°C)') +
  stat_smooth(method = "lm", col = "purple") +
  annotate('text', x = 31, y = 75, label = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))
}


fit1 <- lm(yield ~ tmax, data = sy_temps)
ggplotRegression(fit1)
```

### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.
```{r}
# install and load the packages
library(sf) #df to spatial df
library(USAboundaries) #USA states and counties
library(mapview) #Interactive maps
mapviewOptions(fgb = FALSE)

# using USA boundaries county dataset 
counteez <- us_counties() 
counteez <- counteez[c('state_name', 'countyfp', 'geometry')] %>%
  dplyr::filter(state_name == "Iowa") %>%
  st_transform(2163)

sy_co_temps <- inner_join(counteez, sy_temps,
                          by='countyfp') 

sy_co_temps %>%
  select(c(countyfp, tmax, geometry)) %>%
  arrange(tmax) %>%
  mapview(.,zcol = 'tmax',
         cex = 'tmax', legend = T)
```
**RESPONSE** <br/>
**There appears to be a transition to cooler temperatures from the south west corner of the state to the east and northern boundaries.**

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
```{r}
# using USA boundaries county dataset 
counteez <- us_counties() 
counteez <- counteez[c('state_name', 'countyfp', 'geometry')] %>%
  dplyr::filter(state_name == "Iowa") %>%
  st_transform(2163)

cy_co_temps <- inner_join(counteez, cy_temps,
                          by='countyfp') 


cy_co_temps %>%
  select(c(countyfp, yield, geometry)) %>%
  arrange(yield) %>%
  mapview(.,zcol = 'yield',
         cex = 'yield', legend = T)
```
**RESPONSE** <br/>
**There appears to be a transition to higher yeilds from the southwest corner to the northern and eastern borders of Iowa. This seems to correlate with the negative correlation between summer max temperatures and corn yield visualized above.**
